---
title: "Homework #5"
subtitle: "Exercises from Chapter 4.6: 1 and 2"
description: |
    **Assigned:** Friday, September 29th, 2023 <br>
    **Due:** Friday, October 6th, 2023
author: 
    -   name: Brynjólfur Gauti Guðrúnar Jónsson
        email: brynjolfur@hi.is
        url: https://www.hi.is/staff/brynjolfur
        affiliation: 
        -   name: School of Engineering and Natural Sciences, University of Iceland
            city: Reykjavík
            url: https://english.hi.is/school_of_engineering_and_natural_sciences
# Change this to pdf if you'd rather work in that format
format: html 
editor: source
title-block-banner: true
#### The setting below makes our HTML document self-contained so that we can turn it in ####
#### UNCOMMENT THE LONE BELOW SO THAT YOU CAN TURN IN YOUR HTML DOCUMENT ###################
# embed-resources: true 
---

{{< downloadthis index.qmd dname="hw5" label="Download the Quarto document used to render this file" type=primary class=downloadbutton >}}

# 1

A clinical trial gave six subjects a placebo and six subjects a new weight loss medication. The response variable is the change in weight *(pounds)* from baseline *(so -2.0 means the subject lost 2 pounds)*. The data for the 12 subjects are:

```{r}
#| echo: false
dplyr::tribble(
    ~Placebo, ~Treatment,
    2.0, -3.5,
    -3.1, -1.6,
    -1.0, -4.6,
    0.2, -0.9,
    0.3, -5.1,
    0.4, 0.1
) |> 
    gt::gt()
```

Conduct a Bayesian analysis to compare the means of these two groups. Would you say the treatment is effective? Is your conclusion sensitive to the prior?

---

**Solution:**


---

# 2

Load the classic Boston Housing Data in R:

```{r}
data(Boston, package = "MASS")
```

The response variable is `medv`, the median value of owner-occupied homes *(in $1,000s)*, and the other 13 variables are covariates that describe the neighborhood.


a. Fit a Bayesian linear regression model with uninformative Gaussian priors for the regression coefficients. Verify the MCMC sampler has converged, and summarize the posterior distribution of all regression coefficients.

---

**Solution:**

---

b. Perform a classic least squares *(e.g. using the `lm` function in `R`)*. Compare the results numerically and conceptually with the Bayesian results.



---

**Solution:**


---


c. Refit the Bayesian model with double exponential priors for the regression coefficients, and discuss how the results differ from the analysis with uninformative priors.

---

**Solution:**

---

d. Fit a Bayesian linear regression model in *a.* using only the first 500 observations and compute the posterior predictive distribution for the final 6 observations. Plot the posterior predictive distribution versus the actual value for these 6 observations and comment on whether the predictions are reasonable.

---

**Solution:**

---
