---
title: "Homework #3"
subtitle: Chapters 2.1, 2.2, 2.3 and 2.4
description: |
    **Assigned:** Friday, September 8th, 2023 <br>
    **Due:** Friday, September 15th, 2023
author: 
    -   name: Brynjólfur Gauti Guðrúnar Jónsson
        email: brynjolfur@hi.is
        url: https://www.hi.is/staff/brynjolfur
        affiliation: 
        -   name: School of Engineering and Natural Sciences, University of Iceland
            city: Reykjavík
            url: https://english.hi.is/school_of_engineering_and_natural_sciences
# Change this to pdf if you'd rather work in that format
format: html 
editor: source
theme: flatly
title-block-banner: true
# The setting below makes our HTML document self-contained so that we can turn it in
embed-resources: true 
---

```{r}
#| echo: fenced
#| output: hide

library(tidyr)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(patchwork)
library(invgamma)
theme_set(theme_classic())
```

# Chapter 2.4

## Exercise 6

An assembly line relies on accurate measurements from an image-recognition algorithm at the first stage of the process. It is known that the algorithm is unbiased, so assume that measurements follow a normal distribution with mean zero, 

$$
Y_i\vert \sigma^2\overset{\mathrm{iid}}{\sim}\mathrm{Normal}(0, \sigma^2).
$$

Some errors are permissible, but if $\sigma$ exceeds the threshold $c$ then the algorithm must be replaced. 

You make $n = 20$ measurements and observe 

$$
\sum_{i=1}^n Y_i = -2 \quad \mathrm{and} \quad \sum_{i=1}^n Y_i^2 = 15,
$$ 

and conduct a Bayesian analysis with $\mathrm{InvGamma}(a,b)$ prior. compute the posterior probability that $\sigma > c$ for:

a. $c=1$ and $a=b=0.1$

-----

**Solution:**



-----

b. $c=1$ and $a=b=1.0$

-----

**Solution:** 



-----

c. $c=2$ and $a=b=0.1$

-----

**Solution:**



-----

d. $c=2$ and $a=b=1.0$

-----

**Solution:**




-----

## Exercise 16

Say $Y\vert\lambda \sim \mathrm{Gamma}(1, \lambda).$

a. Derive and plot the Jeffreys' prior for $\lambda$

-----

**Solution:** 

-----

b. Is this prior proper?

-----

**Solution:** This is not a proper prior since it does not integrate to one.

-----

c. Derive the posterior and give conditions on $Y$ to ensure it is proper.

-----

**Solution:**



-----

## Exercise 18

The data in the table below are the result of a survey of commuters in 10 counties likely to be affected by a proposed addition of a high occupancy vehicle (HOV) lane.

```{r}
tibble::tribble(
    ~County, ~Approve, ~Disapprove,
    1, 12, 50,
    2, 90, 150,
    3, 80, 63,
    4, 5, 10,
    5, 63, 63,
    6, 15, 8,
    7, 67, 56,
    8, 22, 19,
    9, 56, 63,
    10, 33, 19
)
```

a. Analyze the data in each county sepparately using the Jeffreys' prior distribution and report the posterior 95% credible set for each county.

-----

**Solution:**

-----

b. Let $\hat p_i$ be the sample proportion of commuters in county $i$ that approve of the HIV lane (e.g. $\hat p_1 = 12/(12+50)=0.194$). Select $a$ abd $b$ so that the mean and variance of the $\mathrm{Beta}(a,b)$ distribution match the mean and variance of the sample proportions $\hat p_1, \dots, \hat p_{10}$.

-----

**Solution:**

-----

c. Conduct an empirical Bayesian analysis by computing the 95% posterior credible sets that results from analyzing each coutny separately using the $\mathrm{Beta}(a,b)$ prior you computed in *b.*

-----

**Solution:**

-----

d. How do the results from *a.* and *c.* differ? What are the advantages and disadvantages of these two analyses?

-----

**Solution:**

-----

